---
title: "Summary of the LIFE child cohort pilot study"
author: "Alexander Hübner"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    pdf_document:
        fig_width: 8
        fig_height: 6
---

```{r libraries, echo=F}
library(knitr)
library(rmarkdown)
library(phyloseq)
library(data.table)
library(tidyverse)
library(pander)
library(patchwork)
library(ggrepel)
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dpi = 150)
theme_set(theme_classic(base_size = 10))
```

```{r load_data}
# Sample overview
samples <- fread("../01-resources/overview_pilot_batch.tsv")
# Sequencing information
nreads <- fread("../05-results/PREP_seqdata_processing_nReads_pilot.tsv")
# Host DNA saliva samples from the Amazonas
amaz_hostdna <- fread("../01-resources/amazonian_saliva_samples_hostDNA.tsv")
# MetaPhlAn4 results
mp4 <- fread("../05-results/COMP_TaxProfiler_MetaPhlAnv4_Jun23_pilot.tsv")
# MetaPhlAn3 results merged with the HMP data set
mp3 <- load("../05-results/COMP_MetaPhlAn3_pilot_HMP2012_phyloSeq.RData")
# Strain analysis
strain <- fread("../05-results/STRA_pilot_strains_sitefreqspectrum.tsv",
                colClasses = c("pattern"="character"))
```

This report summarises the results of the initial sequencing of three supragingival plaque samples
and an unused dental swab (Table 1). The three plaque samples comprise a child from the LIFE
child cohort sample (LLP001.A) and a mother-child pair (LLP002.A and LLP003.A). The second child is
matched in age and sex with the child from the LIFE child cohort.

**Table 1**: **Overview of the samples.**

```{r sample_table, results="asis"}
filter(samples, str_detect(sample_id, "LLP")) |>
select(`Pandora ID` = sample_id,
       `Sample ID` = external_sample_id,
       `additional information` = notes) |>
mutate(`Sample ID` = str_replace(`Sample ID`, " \\(LC[0-9]\\?\\)", "")) |>
(\(x) pandoc.table(x, split.tables = Inf, split.cells = 30))()
```

All samples were extracted with the Qiagen PowerSoil kit and 250 ng of DNA were sheared to a mean
DNA molecule length of 500 bp prior to building Illumina sequencing libraries using an in-house
protocol for double-stranded DNA. The samples were pooled and sequenced across eight lanes of an
Illumina NovaSeq X paired-end sequencing run with nominal length of 2x 150 bp.

## Overview of the sequencing results

After the initial sequencing, adapter sequences were trimmed from the sequencing data and the
trimmed sequences were evaluated for their quality. Low quality reads were discarded (Figure
1a), while the remainder of the sequences were aligned against the telomere-to-telomere assembly
of the human reference genome, _T2T-CHM13v2.0_. 

```{r seqinfo}
seqinfo <- nreads |>
  filter(str_sub(sample, 1, 3) == "LLP") |>
  mutate(host_reads = final_reads - non_host_reads,
         discarded_reads = raw_reads - final_reads) |>
  select(sample, discarded_reads, host_reads, non_host_reads, raw_reads) |>
  pivot_longer(-c(sample, raw_reads), names_to = "read_type", values_to = "nReads") |>
  mutate(frac = nReads / raw_reads,
         read_type = recode(read_type,
                            `discarded_reads` = "discarded",
                            `host_reads` = "host DNA",
                            `non_host_reads` = "other"),
         read_type = factor(read_type, levels = c("discarded", "host DNA", "other")),
         sample = factor(sample, levels = unique(sample)))

# Panel A: overview of fraction of host and non-host DNA
seqinfo_panela <- seqinfo |>
  ggplot(aes(x = fct_rev(sample), y = frac, group = sample)) +
  geom_col(aes(fill = read_type)) +
  coord_flip() +
  labs(y = "fraction of sequencing data") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("#636e72", "#dfe6e9", "#ff7675")) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.title.y = element_blank(),
    axis.line = element_blank(),
    axis.ticks.y = element_blank()
  )

# Panel B: comparison of host DNA fraction to saliva samples from the Amazonas
seqinfo_panelb <- bind_rows(
    amaz_hostdna |>
      mutate(frac = host_reads / total_reads,
             sample_type = "saliva") |>
      select(sample, sample_type, frac),
    seqinfo |>
      filter(read_type == "host DNA",
             sample %in% c("LLP001.A0101", "LLP002.A0101", "LLP003.A0101")) |>
      mutate(sample_type = "plaque") |>
      select(sample, sample_type, frac)
) |>
mutate(sample_type = factor(sample_type, levels = c("plaque", "saliva"))) |>
ggplot(aes(x = sample_type, y = frac, group = sample_type)) +
geom_boxplot(outlier.shape = NA) +
geom_jitter(aes(fill = sample_type), shape = 21, width = .25, size = 2.5) +
labs(y = "fraction of host DNA") +
scale_y_continuous(labels = scales::percent_format(),
                   limits = c(0, 1)) +
theme(
  legend.position = "none",
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.line = element_blank(),
  panel.grid.major.y = element_line(linewidth = 0.5,
        colour = "grey50",
        linetype = 3)
    )

seqinfo_panela + seqinfo_panelb +
plot_annotation(tag_levels = "a")
```

**Figure 1**: **Overview of the origin of the sequencing data**. **a** Fraction of reads that were
either discarded due to low quality or were identified as host DNA. **b** Comparison of the fraction
of host DNA between the plaque samples and saliva samples from individuals from the Amazonas region
in South America.

From the four samples, the blank dental swap (_LLP004.A0101_) had the highest fraction of discarded
sequences (Figure 1a). This is expected because, due to its low biomass, a higher fraction of
sequencing artefacts, such as adapter dimers, is generated during the library preparation. For the
actual plaque samples, the fraction of discarded sequences is about 10% - this order of magnitude is
regularly observed in other biological samples, too.

The amount of host DNA varied quite substantially across the plaque samples spanning from 18% to
74% (Figure 1a). These values are quite high compared to matured oral biofilm. However, when
comparing these samples against saliva samples from individuals from the Amazonas basin in South
America (Figure 1b), the plaque samples fell within the lower end of the fraction of host DNA
distribution of the saliva samples. Only sample _LLP003.A0101_ fell within the interquartile
distribution of the saliva samples indicating that this sample had a particularly high proportion of
host DNA.

As there is no ethical approval to study the host DNA, it would make sense to think about laboratory
strategies to reduce the fraction of host DNA prior to building the sequencing libraries, e.g. by
removing the host DNA via a spin column. The amount of sequencing data available for
microbiome-related analyses was higher than 50 million sequences, which is commonly found in other
microbial studies, for two out of three samples (Table 2).

**Table 2**: **Number of sequences available for microbial DNA analyses.**

```{r tbl_nonhost, results="asis"}
seqinfo |>
filter(read_type == "other", sample != "LLP004.A0101") |>
mutate(`non-host sequences [million]` = nReads / 1e6) |>
select(sample, `non-host sequences [million]`) |>
(\(x) pandoc.table(x, split.tables = Inf,
                   round = 1,
                   keep.trailing.zeros = TRUE))()
```

## Analysis of the microbial DNA

After discarding the host DNA, I inferred which microbial species are present in the plaque samples
and laboratory controls by taxonomically profiling it with the program MetaPhlAn. MetaPhlAn uses a
catalogue of marker genes that are identified from many microbial species that are present in
human-associated microbiome samples. Because it uses a subset of high-quality microbial genomes that
are human-associated, it reduces the risk for false positive and false negative species assignment
and returns accurate taxonomic profiles.

First, I investigated how many microbial species were present in each sample (Figure 2).

```{r alpha_diversity, fig.height=3}
mp4_lt <- mp4 |>
  pivot_longer(-clade_name, names_to = "sample", values_to = "relab") |>
  mutate(taxon = str_extract(clade_name, "s__.+$"))

n_obs_species <- mp4_lt |>
  select(-clade_name) |>
  summarise(n = sum(relab > 0), .by = sample) |>
  mutate(sample_type = case_when(
    str_sub(sample, 1, 3) == "EXB" ~ "extraction blank",
    str_sub(sample, 1, 3) == "LIB" ~ "library blank",
    sample == "LLP004.A0101" ~ "swab blank",
    .default = "plaque"
    ))

n_obs_species |>
mutate(sample_type = factor(sample_type, levels = c("plaque", "swab blank",
                                                    "extraction blank",
                                                    "library blank"))) |>
ggplot(aes(x = 1, y = n)) +
geom_jitter(aes(fill = sample_type), shape = 21, size = 3, width = .25) +
geom_hline(yintercept = 50, linetype = 1, linewidth = 0.5, colour = "red") +
geom_text_repel(data = filter(n_obs_species,
                              sample %in% c("LLP001.A0101", "LLP002.A0101", "LLP003.A0101")),
               aes(label = sample)) +
coord_flip() +
labs(y = "number of observed species") +
theme(
  legend.position = "top",
  legend.title = element_blank(),
  axis.title.y = element_blank(),
  axis.text.y = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.grid.major.x = element_line(colour = "grey50",
                                    linetype = 3, linewidth = 0.5)
)
```

**Figure 2**: **The number of observed taxa inferred from MetaPhlAn v4.**

Across the samples, the three plaque samples had by far the highest number of observed species,
while the blank samples had much lower numbers (Figure 2). This is expected as blank samples
should have a low biomass compared to the true biological samples. Among the three samples, the
sample with the highest number of non-host DNA (Figure 1a), _LLP001.A0101_, also had the highest
number of observed species, however, this trend was not confirmed by the other two plaque samples.

Next to having a higher number of species, the most abundant species of the plaque samples strongly
differed from the species that were most abundant in the blank samples (Figure 3). Only a single
species, _Neisseria sicca_, which was present in a single extraction blank sample, was shared within
the top 20 most abundant species between the plaque samples and the blank samples (Figure 3a).

```{r most_abundant_species}
mas_per_type <- mp4_lt |>
  mutate(sampletype = if_else(str_sub(sample, 1, 6) %in% c("LLP001", "LLP002", "LLP003"),
                              "plaque", "blank")) |>
  summarise(meanRelAb = mean(relab),
            .by = c(sampletype, taxon))
mas_plaque <- mas_per_type |>
  filter(sampletype == "plaque") |>
  arrange(desc(meanRelAb)) |>
  head(20) |>
  pull(taxon)
mas_blank <- mas_per_type |>
  filter(sampletype == "blank", taxon != "s__Neisseria_sicca") |>
  arrange(desc(meanRelAb)) |>
  head(20) |>
  pull(taxon)

mas_panel_a <- mp4_lt |>
  filter(taxon %in% mas_plaque) |>
  mutate(relab = relab / 100,
         family = str_extract(clade_name, "f__[A-Za-z0-9]+"),
         taxon = factor(taxon, levels = mas_plaque),
         sampletype = if_else(str_sub(sample, 1, 6) %in% c("LLP001", "LLP002", "LLP003"),
                              "plaque", "blank"),
         sampletype = factor(sampletype, levels = c("blank", "plaque"))
      ) |>
  arrange(sampletype, desc(sample)) |>
  (\(x) mutate(x, sample = factor(sample, levels = unique(x$sample))))() |>
  arrange(taxon) |>
  ggplot(aes(x = sample, y = relab, group = sample)) +
  geom_col(aes(fill = taxon)) +
  coord_flip() +
  labs(y = "relative abundance",
       fill = "top 20 plaque species") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(
    legend.text = element_text(size = 8),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.5, linetype = 3, colour = "grey50")
      ) +
  guides(fill = guide_legend(ncol = 2))

mas_panel_b <- mp4_lt |>
  filter(taxon %in% mas_blank) |>
  mutate(relab = relab / 100,
         family = str_extract(clade_name, "f__[A-Za-z0-9]+"),
         taxon = factor(taxon, levels = mas_blank),
         sampletype = if_else(str_sub(sample, 1, 6) %in% c("LLP001", "LLP002", "LLP003"),
                              "plaque", "blank"),
         sampletype = factor(sampletype, levels = c("blank", "plaque"))
      ) |>
  arrange(sampletype, desc(sample)) |>
  (\(x) mutate(x, sample = factor(sample, levels = unique(x$sample))))() |>
  arrange(taxon) |>
  ggplot(aes(x = sample, y = relab, group = sample)) +
  geom_col(aes(fill = taxon)) +
  coord_flip() +
  labs(y = "relative abundance",
       fill = "top 20 blank species") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(
    legend.text = element_text(size = 8),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.5, linetype = 3, colour = "grey50")
      ) +
  guides(fill = guide_legend(ncol = 2))

mas_panel_a + mas_panel_b +
plot_layout(nrow = 2) +
plot_annotation(tag_levels = "a")
```

**Figure 3**: **Overview of the top 20 most abundant species that were either found in the plaque or
the blank samples.**

The top 20 most abundant species in the plaque samples are strongly dominated by species associated
with human oral microbiome (Figure 3a). In contrast, the blank swab sample, _LLP004.A0101_, was
dominated by _Brucella intermedia_ and _Burkholderia multivorans_ (Figure 3b). Neither species
is associated with the human oral microbiome but have been isolated as opportunistic pathogens,
suggesting that these are primarily found in the environment. For the remaining extraction and
library blank samples, some of these were contaminated with bacteria associated with the human gut
microbiome. As the plaque samples were co-extracted alongside FTA cards containing human stool
samples, this is expected.

Although these results combined strongly support the plaque samples were true oral microbiome
samples, there is still the open question whether our wet lab analysis protocols would have
introduced a systematic bias compared to the microbial profiles of published oral microbiome
samples. To answer this question, I compared the taxonomic profiles of the three plaque samples and
the empty swab to the profiles of the human oral microbiome samples of the Human Microbiome Project
(HMP; Figure 4). I compared the plaque samples against four different sample types available in
the Human Microbiome Project: supragingival plaque, subgingival plaque, buccal mucosa, and tongue
dorsum.

```{r pcoa}
phy <- prune_samples(as.logical(!(as(sample_data(phy), "data.frame")$sample_type %in%
                                  c("extraction blank", "library blank"))), phy)
sample_data(phy) <- as(sample_data(phy), "data.frame") |>
  mutate(sample_type = if_else(sample == "LLP004.A0101", "blank swab", sample_type),
         sample_type = factor(sample_type, levels = c("LIFE Child cohort plaque",
                                                      "HMP supragingival plaque",
                                                      "HMP subgingival plaque",
                                                      "HMP buccal mucosa",
                                                      "HMP tongue dorsum",
                                                      "blank swab"))) |>
  sample_data()
  
bray_dist <- ordinate(phy, "PCoA", "bray")

ord_df <- bray_dist$vectors[, 1:2] |>
  as.data.frame() |>
  rownames_to_column(var = "sample") |>
  rename(PC1 = Axis.1, PC2 = Axis.2) |>
  as_tibble() |>
  left_join(as(sample_data(phy), "data.frame") |>
            select(sample, sample_type),
            by = "sample") |>
  mutate(study = if_else(sample_type == "LIFE Child cohort plaque", "this study", "HMP"))

ord_df |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = sample_type, size = study, alpha = study),
                 shape = 21) +
  labs(x = str_c("PC1", ": ", round(bray_dist$values$Eigenvalues[1], 1), "%"),
       y = str_c("PC2", ": ", round(bray_dist$values$Eigenvalues[2], 1), "%"),
       fill = "") +
  scale_fill_manual(values = c("#c23616", "#e1b12c", "#22a6b3", "#6ab04c", "#7f8fa6", "#e056fd")) +
  scale_size_manual(values = c(2, 4)) +
  scale_alpha_manual(values = c(0.8, 1)) +
  theme(
    legend.position = "top"
  ) +
  guides(fill = guide_legend(override.aes = list(pch = 21, size = 4), nrow = 2),
         alpha = "none",
         size = "none")
```

**Figure 4**: **Principal component analysis of the LIFE Child cohort plaque samples with the HMP
oral microbiome samples.** The underlying dissimilarity metric is Bray-Curtis.

For the comparison, I performed a beta diversity analysis based on Bray-Curtis dissimilarity. Each
pair of samples is compared and their difference in species composition is recorded. The matrix of
pairwise dissimilarity scores is then used as the input in a Principal Coordinate Analysis (PCoA).

The three plaque samples clustered with the other supragingival and subgingival plaque samples from
the HMP, highlighting that our wet lab protocol did not introduce a systematic bias. This cluster of
plaque samples was distinct from the cluster of buccal mucosa and the cluster of tongue dorsum
samples. The blank swab sample did not cluster with any microbiome sample but fell in the origin of
the PCoA plot.

In summary, this result highlighted that the applied wet lab protocols did not introduce a
systematic bias but that we generated high-quality microbial sequencing data that allows us to
compare the plaque samples with each other but also with published oral microbiome samples.

## Strain sharing between individuals

Our sample set of three comprises a mother-child pair and one individual child. Such a setup allows
asking whether the strains observed in the mother-child pair are genetically more closely to each
other than they are to the unrelated child? Previous work on strain sharing in the oral microbiome
has shown that although not very microbial species is vertically transmitted, there is plenty of
evidence for it.

The ability of detecting strain sharing is dependent on whether there are sufficient amounts of
genetic information in each sample. In the case that we are not able to reconstruct the genome for
one of the three samples, then we are also not able to investigate the strain sharing across the
samples. Therefore, I calculated the median relative abundance of all microbial species across the
three plaque samples and selected the top three species: _Neisseria sicca_, _Lautropia mirabilis_,
and _Streptococcus oralis_. 

I downloaded the species representative genome for each species from NCBI Genomes and aligned the
short-read sequencing data of each sample against the reference genomes. After the sequence
alignment, I jointly called genotypes across the three samples for each species using freeBayes. I
filtered the inferred genotypes and only kept them when the genotype quality was greater than 30,
there was genotype information for all samples, and when the site had only two alleles (bi-allelic;
**Table 3).

**Table 3**: **The number of sites per species for which a pair of samples share a variant.**

```{r strain_analysis, results="asis"}
n_snps <- strain |>
  summarise(total = sum(count), .by = species)

strain |>
mutate(sharers = case_when(
         pattern %in% c("001", "110") ~ "LLP002 & LLP003",
         pattern %in% c("011", "100") ~ "LLP001 & LLP002",
         pattern %in% c("010", "101") ~ "LLP001 & LLP003",
        .default = "non\ninformative"
    ),
    species = recode(species, `Nsicca` = "N. sicca",
                     `Lmirabilis` = "L. mirabilis",
                     `Soralis` = "S. oralis")) |>
summarise(count = sum(count), .by = c(species, sharers)) |>
filter(sharers != "non\ninformative") |>
pivot_wider(names_from = "sharers", values_from = "count") |>
(\(x) pandoc.table(x, big.mark = ",", split.tables = Inf))()
```

This filtering scheme left me with between `r format(min(n_snps$total), big.mark = ",")` and 
`r format(max(n_snps$total), big.mark = ",")` variants for each species. I grouped the variants in
three bins dependent which two samples shared a variant and were different from the third sample
(Table 3).

From the three species, we only observed an excess of shared variants between the mother (LLP002)
and her child (LLP003) for _N. sicca_. Here, the pair shared three times more variants with each
other than either of them shared with the unrelated child (LLP001). In contrast, for the species
_L. mirabilis_ and _S. oralis_, the two children shared many more variants with each other than the
mother-child pair. This highlights the reported issues of vertical transmission of strains from
mother to their children and its stability through the lifetime of their offspring.

## Conclusion

In this pilot experiment, we extracted DNA from three supragingival plaque samples, converted it
into sequencing libraries and sequenced them deeply. After initial processing for sequence quality,
we obtained between 34 and 173 million microbial sequences for the plaque samples. The huge
variation across the plaque samples can be explained by the difference in the fraction of host DNA,
and it might be worth to consider methods to reduce the amount of host DNA prior to the sequencing.

Taxonomic profiling of the plaque samples showed that the plaque samples had many more microbial
species than the accompanying blanks and that the species strongly differed between them. A
comparison of taxonomic profiles of the plaque samples to previously published human oral microbiome
samples from the Human Microbiome Project revealed that the plaque samples clustered with other
plaque samples distinctly from other oral cavity samples, such as buccal mucosa or tongue dorsum.
Therefore, the applied wet lab protocols did not introduce a systematic bias compared to previously
published samples.

An initial investigation on whether strain sharing between the mother-child pair was frequently
observed revealed that only for one of the three tested species, _N. sicca_, there was an excess of
variants shared between the mother-child pair. For the species _L. mirabilis_ and _S. oralis_, the
two children shared more variants with each other than the mother and her offspring.
